Periodograms for Multiband Astronomical Time Series

### 2015-08-21 writeup/paper.tex
more tweaks in response to second referee

The most popular non-parametric method is the phase dispersion minimization (PDM) introduced by \cite{PDM1978}. 
Dispersion per bin is computed for binned phased light curves evaluated for a grid of trial periods. The best
period minimizes the dispersion per bin.  A similar and related non-parametric method that has been recently 
gaining popularity is the Supersmoother routine \citep{Reimann94}. It uses a running mean or running linear 
regression on the data to fit the observations as a function of phase to a range of periods. The best period 
minimizes a figure-of-merit, adopted as weighted sum of absolute residuals around the running mean. 
Neither the Supersmoother algorithm nor the PDM method require \apriori{} knowledge of the light curve shape. 
We note that neither method produces posterior probability distribution for the period but rather a single point 
estimate.

The most popular parametric method is the Lomb-Scargle periodogram, which is discussed in detail in \sect{brief_overview}.
The Lomb-Scargle periodogram is related to the $\chi^2$ for a least-square fit of a single sinusoid to data
and can treat non-uniformly sampled time series with heteroscedastic measurement uncertainties. 
\old{The underlying model of the Lomb–Scargle periodogram is nonlinear in frequency and basis functions at different frequencies are not orthogonal. 
As a result, the periodogram has many local maxima.}
\new{The underlying model of the Lomb-Scargle periodogram is nonlinear in frequency, and thus the likelihood surface in frequency is non-convex.
This non-convexity is readily apparent in the many local maxima of the typical periodogram, which makes it difficult to find the maximum via standard numerical optimization routines.}
Thus in practice the global 
maximum of the periodogram is often found by a brute-force grid search \citep[for details see, e.g.][]{ICVG2014}.
A more general parametric method based on the use of continuous-time autoregressive moving average (CARMA) model
was recently introduced by \citet{Kelly14}. CARMA models can also treat non-uniformly sampled time series with 
heteroscedastic measurement uncertainties, and can handle complex variability patterns.

A weakness of the above methods is that they require homogeneous measurements -- for astronomy data, this means 
that successive measurements must be taken through a single photometric bandpass (filter). This has not been a major
problem for past surveys because measurements are generally taken through a single photometric filter 
\citep [e.g. LINEAR,][]{LINEAR1}, or nearly-simultaneously in all bands at each observation \citep [e.g. SDSS,][]{Sesar2010}.
For the case of simultaneously taken multiband measurements, \cite{Suveges12} utilized the principal component
method to optimally extract the best period. Their method is essentially a multiband generalization of the well-known
two-band Welch-Stetson variability index \citep{WelchStetson1993}. Unfortunately, when data in each band are taken at
different times, such an approach in not applicable. In such cases, past studies have generally relied 
on \adhoc{} methods such as a majority vote among multiple single-band estimates of the 
periodogram \citep[e.g.,][]{Oluseyi12}.

@@

\section{Standard Least Squares Spectral Fitting}
\sectlabel{standard_least_squares}
In this section we present a brief quantitative introduction to the least squares fitting formulation of the normalized periodogram of \eq{LombScargle}. We denote $N$ observed data points as
\begin{equation}
  D = \{t_k, y_k, \sigma_k\}_{k=1}^N
\end{equation}
where $t_k$ is the time of observation, $y_k$ is the observed value (typically a magnitude), and $\sigma_k$ describes the Gaussian errors on each value. For notational simplicity we will assume without loss of generality that the data $y_k$ are centered such that the measurements within each band satisfy
\begin{equation}
  \eqlabel{ycentered}
  \frac{\sum_k w_ky_k}{\sum_k w_k} = 0
\end{equation}
where the weights are $w_k = \sigma_k^{-2}$.
Though this assumption is essential to the models presented in this section, it will become superfluous with the floating-mean models described in later sections.

@@

As an example of one of these generalizations, we'll consider 
what has variously been called the {\it Date-compensated Discrete Fourier Transform} \citep{Ferraz-Mello81}, the {\it floating-mean periodogram} \citep{Cumming99}, and the {\it generalized Lomb-Scargle method} \citep{Zechmeister09}. Here we use the term {\it floating-mean periodogram}. This method adjusts the classic normalized periodogram by fitting the mean of the model alongside the amplitudes:
\begin{equation}
  y(t~|~\omega, \theta) = \theta_0 + \theta_1\sin\omega t + \theta_2\cos\omega t
\end{equation}
The periodogram derived from this model can be more accurate than the standard periodogram for certain observing cadences and selection functions. \citet{Zechmeister09} detail the required modifications to the orthogonalized harmonic formalism of \eq{LombScargle} to allow the mean to float in the model. In the matrix formalism, the modification is much more straightforward: all that is required is to add a column of ones to the $X_\omega$ matrix before computing the power via \eq{LombScargle2}. This column of ones corresponds to a third entry in the parameter vector $\theta$, and acts as a uniform constant offset for all data points.

@@

\new{One might hope that when adding terms, the correct-period model would show more of an improvement than the incorrect-period model (and thus the periodogram maximum would become more pronounced), but this does not generally hold.
Consider that in the extreme limit in which the number model parameters is equal to the number of data points, the model has enough flexibility to fit the data perfectly at {\it every} frequency, and the resulting periodogram would be a flat line at maximum power!
This can only be the case if, on average, addition of terms preferentially boosts the background level.}

@@

The previous sections raise the question: how complicated a model should we use? We have seen that as we add more terms to the fit, the model will come closer and closer to the observed data.  In the extreme case, when the number of model parameters equals the number of data points, the model can fit the data {\it exactly} regardless of frequency and the periodogram will be everywhere unity (though in most cases, numerical inaccuracies prevent a truly perfect fit). For very high-order models, the fit becomes very sensitive to the noise in the data, and we say we have {\it over-fit} the data. This can be addressed by explicitly truncating the series, but we can also use a {\it regularization} term to mathematically enforce a less complicated model.

@@

\subsection{Relationship of Multiband and Single-band approaches}
\sectlabel{relationship}
\new{
With this formalism in place, we can return to the na{\"i}ve models discussed at the beginning of \sect{multiband}.
The first is simply a standard Lomb-Scargle, ignoring the band information.
The second, in which each band is fit independently, turns out to be equivalent to the $(N_{base},N_{band})=(0,1)$ model defined above.
}
\old{
The multi-phase model with $(N_{base},N_{band})=(0,1)$ turns out to be a particularly special case.
}
Here the base model is a simple global offset which is degenerate with the offsets in each band, so that the design matrix $X_\omega$ can be straightforwardly rearranged as block-diagonal.
A block-diagonal design matrix in a linear model indicates that components of the model are being solved independently: here these independent components amount to the single-band floating-mean model from \sect{floating_mean}, fit independently for each of the $K$ bands.
\old{This particular multiband model can give us insight into the relationship between single-band and multiband approaches.}

For band $k$, we'll denote the single-band floating-mean periodogram as
\begin{equation}
  P_N^{(k)}(\omega) = 1 - \frac{\chi^2_{min, k}(\omega)}{\chi^2_{0,k}}
\end{equation}
The full multiband periodogram is given by
\begin{equation}
  P_N^{(0,1)}(\omega) = 1 - \frac{\sum_{k=1}^K\chi^2_{min, k}(\omega)}{\sum_{k=1}^K\chi^2_{0,k}}
\end{equation}
and it can easily be shown that the $P_N$ can be constructed as a weighted sum of $P_N^{(k)}$:
\begin{equation}
  P_N^{(0,1)}(\omega) = \frac{\sum_{k=1}^K\chi^2_{0,k}P_N^{(k)}}{\sum_{k=1}^K\chi^2_{0,k}}.
\end{equation}
Thus the $(N_{base},N_{band})=(0,1)$  multiband periodogram is identical to a weighted sum of standard periodograms in each band, where the weights $\chi^2_{0,k}$ are a reflection of both the number of measurements in each band and how much those measurements deviate from a simple constant reference model.

@@

The full S10 RR Lyrae dataset consists of 483 objects with an average of 55 observations in each of the five SDSS $ugriz$ bands spread over just under ten years.  In the upper panels of \fig{compare_periods} we show the observed data for one of these objects, along with the periodogram derived with the single-band supersmoother model and the shared-phase $(0, 1)$-multiband model\footnote{The supersmoother ``periodogram'' $P_{SS}$ is constructed from the minimum sum of weighted model residuals $\bar{r}_{min}$ in analogy with \eq{chi2PN}: $P_{SS}(\omega) = 1 - \bar{r}_{min}(\omega) / \bar{r}_0$, where $\bar{r}_0$ is the mean absolute residual around a constant model.}. Here we have a case which is analogous to that shown for simulated data in the top panels of \fig{multiband_sim}: each band has enough data to easily locate candidate peaks, the best of which is selected via the S10 template-fitting procedure.

>>>>

The most popular non-parametric method is the phase dispersion minimization (PDM) introduced by \cite{PDM1978}. 
Dispersion per bin is computed for binned phased light curves evaluated for a grid of trial periods. The best
period minimizes the dispersion per bin.  A similar and related non-parametric method that has been recently 
gaining popularity is the Supersmoother routine \citep{Reimann94}. It uses a running mean or running linear 
regression on the data to fit the observations as a function of phase to a range of periods. The best period 
minimizes a figure-of-merit, adopted as weighted sum of absolute residuals around the running mean. 
Neither the Supersmoother algorithm nor the PDM method require \apriori{} knowledge of the light curve shape.

The most popular parametric method is the Lomb-Scargle periodogram, which is discussed in detail in \sect{brief_overview}.
The Lomb-Scargle periodogram is related to the $\chi^2$ for a least-square fit of a single sinusoid to data
and can treat non-uniformly sampled time series with heteroscedastic measurement uncertainties. 
The underlying model of the Lomb–Scargle periodogram is nonlinear in frequency \old{and basis functions at different frequencies are not orthogonal. 
As a result, the periodogram has many local maxima.}
\new{and so the likelihood surface in frequency is non-convex.
This non-convexity is readily apparent in the many local maxima of the typical periodogram, which makes it difficult to find the maximum via standard numerical optimization routines.}
Thus in practice the global 
maximum of the periodogram is often found by a brute-force grid search \citep[for details see, e.g.][]{ICVG2014}.

A more general parametric method based on the use of continuous-time autoregressive moving average (CARMA) model
was recently introduced by \citet{Kelly14}. CARMA models can also treat non-uniformly sampled time series with 
heteroscedastic measurement uncertainties, and can handle complex variability patterns.

A weakness of all these standard methods is that they require homogeneous measurements -- for astronomy data, this means 
that successive measurements must be taken through a single photometric bandpass (filter). This has not been a major
problem for past surveys because measurements are generally taken through a single photometric filter 
\citep [e.g. LINEAR,][]{LINEAR1}, or nearly-simultaneously in all bands at each observation \citep [e.g. SDSS,][]{Sesar2010}.
For the case of simultaneously taken multiband measurements, \cite{Suveges12} utilized the principal component
method to optimally extract the best period. Their method is essentially a multiband generalization of the well-known
two-band Welch-Stetson variability index \citep{WelchStetson1993}. Unfortunately, when data in each band are taken at
different times, such an approach in not applicable. In such cases, past studies have generally relied 
on \adhoc{} methods such as a majority vote among multiple single-band estimates of the 
periodogram \citep[e.g.,][]{Oluseyi12}.

@@

\section{Standard Least Squares Spectral Fitting}
\sectlabel{standard_least_squares}
In this section we present a brief quantitative introduction to the least squares fitting formulation of the normalized periodogram of \eq{LombScargle}. We denote $N$ observed data points as
\begin{equation}
  D = \{t_k, y_k, \sigma_k\}_{k=1}^N
\end{equation}
where $t_k$ is the time of observation, $y_k$ is the observed value (typically a magnitude), and $\sigma_k$ describes the Gaussian errors on each value. For notational simplicity we will assume without loss of generality that the data $y_k$ are centered such that the measurements within each band satisfy
\begin{equation}
  \eqlabel{ycentered}
  \frac{\sum_k w_ky_k}{\sum_k w_k} = 0
\end{equation}
where the weights are $w_k = \sigma_k^{-2}$.
Though this assumption is essential to the simpler models presented in this section, it will become superfluous with the floating-mean models described in later sections.

@@

As an example of one of these generalizations, we'll consider 
what has variously been called the {\it Date-compensated Discrete Fourier Transform} \citep{Ferraz-Mello81}, the {\it floating-mean periodogram} \citep{Cumming99}, and the {\it generalized Lomb-Scargle method} \citep{Zechmeister09}. Here we use the term {\it floating-mean periodogram}. This method adjusts the classic normalized periodogram by fitting the mean of the model alongside the amplitudes:
\begin{equation}
  y(t~|~\omega, \theta) = \theta_0 + \theta_1\sin\omega t + \theta_2\cos\omega t
\end{equation}
The periodogram derived from this model can be more accurate than the standard \new{pre-centered} periodogram for certain observing cadences and selection functions, \new{and espeically when searching for long-period varaibility or working with very few samples \citep{Cumming99}.} \citet{Zechmeister09} detail the required modifications to the orthogonalized harmonic formalism of \eq{LombScargle} to allow the mean to float in the model. In the matrix formalism, the modification is much more straightforward: all that is required is to add a column of ones to the $X_\omega$ matrix before computing the power via \eq{LombScargle2}. This column of ones corresponds to a third entry in the parameter vector $\theta$, and acts as a uniform constant offset for all data points.

@@

\new{One might hope that when adding terms, the correct-period model would show more of an improvement than the incorrect-period model (and thus the periodogram maximum would become more pronounced in comparison to the background), but this does not generally hold.
Consider that in the extreme limit in which the number model parameters is equal to the number of data points, the model has enough flexibility to fit the data perfectly at {\it every} frequency, and the resulting periodogram would be everywhere unity!
This can only be the case if, on average, addition of terms preferentially boosts the background level.}

@@

The previous sections raise the question: how complicated a model should we use? We have seen that as we add more terms to the fit, the model will more closely describe the observed data.
\old{In the extreme case, when the number of model parameters equals the number of data points, the model can fit the data {\it exactly} regardless of frequency and the periodogram will be everywhere unity (though in most cases, numerical inaccuracies prevent a truly perfect fit).}
For very high-order models, however, such a close fit {\it over-fits} the data: that is, the fit is more responsive to statistical noise in the observations than to the underlying signal.
This can be addressed by explicitly truncating the series at some number of terms, but we can also use a {\it regularization} term to mathematically enforce model simplicity.

@@

\subsection{Relationship of Multiband and Single-band approaches}
\sectlabel{relationship}
\new{
With this formalism in place, we can return briefly to the na{\"i}ve models discussed at the beginning of \sect{multiband}.
The first, which ignores band information, is simply a standard Lomb-Scargle over the heterogeneous data.
The second, in which each band is fit independently, turns out to be equivalent to the $(N_{base},N_{band})=(0,1)$ model defined above.
}
\old{
The multi-phase model with $(N_{base},N_{band})=(0,1)$ turns out to be a particularly special case.
}
Here the base model is a simple global offset which is degenerate with the offsets in each band, so that the design matrix $X_\omega$ can be straightforwardly rearranged as block-diagonal.
A block-diagonal design matrix in a linear model indicates that components of the model are being solved independently: here these independent components amount to the single-band floating-mean model from \sect{floating_mean}, fit independently for each of the $K$ bands.
\old{This particular multiband model can give us insight into the relationship between single-band and multiband approaches.}

For band $k$, we'll denote the single-band floating-mean periodogram as
\begin{equation}
  P_N^{(k)}(\omega) = 1 - \frac{\chi^2_{min, k}(\omega)}{\chi^2_{0,k}}
\end{equation}
The full multiband periodogram is given by
\begin{equation}
  P_N^{(0,1)}(\omega) = 1 - \frac{\sum_{k=1}^K\chi^2_{min, k}(\omega)}{\sum_{k=1}^K\chi^2_{0,k}}
\end{equation}
and it can be shown straightforwaredly that $P_N^{(0,1)}$ can be constructed as a weighted sum of $P_N^{(k)}$:
\begin{equation}
  P_N^{(0,1)}(\omega) = \frac{\sum_{k=1}^K\chi^2_{0,k}P_N^{(k)}}{\sum_{k=1}^K\chi^2_{0,k}}.
\end{equation}
Thus the $(N_{base},N_{band})=(0,1)$  multiband periodogram is identical to a weighted sum of standard periodograms in each band, where the weights $\chi^2_{0,k}$ are a reflection of both the number of measurements in each band and how much those measurements deviate from a simple constant reference model.

@@

The full S10 RR Lyrae dataset consists of 483 objects with an average of 55 observations in each of the five SDSS $ugriz$ bands spread over just under ten years.
In the upper panels of \fig{compare_periods} we show the observed data for one of these objects, along with the periodogram derived with the single-band supersmoother model\footnote{The supersmoother ``periodogram'' $P_{SS}$ is constructed from the minimum sum of weighted model residuals $\bar{r}_{min}$ in analogy with \eq{chi2PN}: $P_{SS}(\omega) = 1 - \bar{r}_{min}(\omega) / \bar{r}_0$, where $\bar{r}_0$ is the mean absolute residual around a constant model.} and the shared-phase $(0, 1)$-multiband model.
Here we have a case which is analogous to that shown for simulated data in the top panels of \fig{multiband_sim}: each band has enough data to easily locate candidate peaks, the best of which is selected via the S10 template-fitting procedure.
